# $k$-Nearest Neighbors {#knn-reg}

## $k$-Nearest Neighbors

$$
\hat{f}(x) = \frac{1}{k} \sum_{i \in \mathcal{N}_k(x, \mathcal{D})} y_i
$$

## KNN in `R`

```{r}
library(FNN)
library(MASS)
data(Boston)
```

```{r}
set.seed(42)
boston_idx = sample(1:nrow(Boston), size = 250)
trn_boston = Boston[boston_idx, ]
tst_boston  = Boston[-boston_idx, ]
```

```{r}
X_trn_boston = trn_boston["lstat"]
X_tst_boston = tst_boston["lstat"]
y_trn_boston = trn_boston["medv"]
y_tst_boston = tst_boston["medv"]
```

We create an additional "test" set `lstat_grid`, that is a grid of `lstat` values at which we will predict `medv` in order to create graphics.

```{r}
X_trn_boston_min = min(X_trn_boston)
X_trn_boston_max = max(X_trn_boston)
lstat_grid = data.frame(lstat = seq(X_trn_boston_min, X_trn_boston_max, 
                                    by = 0.01))
```

To perform KNN for regression, we will need `knn.reg()` from the `FNN` package. Notice that, we do **not** load this package, but instead use `FNN::knn.reg` to access the function. Note that, in the future, we'll need to be careful about loading the `FNN` package as it also contains a function called `knn`. This function also appears in the `class` package which we will likely use later.

```{r}
pred_001 = knn.reg(train = X_trn_boston, test = lstat_grid, y = y_trn_boston, k = 1)
pred_005 = knn.reg(train = X_trn_boston, test = lstat_grid, y = y_trn_boston, k = 5)
pred_010 = knn.reg(train = X_trn_boston, test = lstat_grid, y = y_trn_boston, k = 10)
pred_050 = knn.reg(train = X_trn_boston, test = lstat_grid, y = y_trn_boston, k = 50)
pred_100 = knn.reg(train = X_trn_boston, test = lstat_grid, y = y_trn_boston, k = 100)
pred_250 = knn.reg(train = X_trn_boston, test = lstat_grid, y = y_trn_boston, k = 250)
```

We make predictions for various values of `k`. Note that `250` is the total number of observations in this training dataset.

```{r, fig.height = 8, fig.width = 6, echo = FALSE}
par(mfrow = c(3, 2))

plot(medv ~ lstat, data = trn_boston, cex = .8, col = "dodgerblue", main = "k = 1")
lines(lstat_grid$lstat, pred_001$pred, col = "darkorange", lwd = 0.25)

plot(medv ~ lstat, data = trn_boston, cex = .8, col = "dodgerblue", main = "k = 5")
lines(lstat_grid$lstat, pred_005$pred, col = "darkorange", lwd = 0.75)

plot(medv ~ lstat, data = trn_boston, cex = .8, col = "dodgerblue", main = "k = 10")
lines(lstat_grid$lstat, pred_010$pred, col = "darkorange", lwd = 1)

plot(medv ~ lstat, data = trn_boston, cex = .8, col = "dodgerblue", main = "k = 25")
lines(lstat_grid$lstat, pred_050$pred, col = "darkorange", lwd = 1.5)

plot(medv ~ lstat, data = trn_boston, cex = .8, col = "dodgerblue", main = "k = 50")
lines(lstat_grid$lstat, pred_100$pred, col = "darkorange", lwd = 2)

plot(medv ~ lstat, data = trn_boston, cex = .8, col = "dodgerblue", main = "k = 250")
lines(lstat_grid$lstat, pred_250$pred, col = "darkorange", lwd = 2)
```

We see that `k = 1` is clearly overfitting, as `k = 1` is a very complex, highly variable model. Conversely, `k = 250` is clearly underfitting the data, as `k = 250` is a very simple, low variance model. In fact, here it is predicting a simple average of all the data at each point.


## Choosing $k$

```{r}
# calculate train RMSE
# calculate test RMSE
```


## Scaling Data

```{r}
sim_knn_data = function(n_obs = 50) {
  x1 = seq(0, 10, length.out = n_obs)
  x2 = runif(n = n_obs, min = 0, max = 2)
  x3 = runif(n = n_obs, min = 0, max = 1)
  x4 = runif(n = n_obs, min = 0, max = 5)
  x5 = runif(n = n_obs, min = 0, max = 5)
  y = x1 ^ 2 + rnorm(n = n_obs)
  data.frame(y, x1, x2, x3,x4, x5)
}
```

```{r}
set.seed(42)
knn_data = sim_knn_data()
```

```{r, echo = FALSE}
par(mfrow = c(1, 2))
orange_blue = c(rep("grey", 45), rep("darkorange", 5))
plot(x1 ~ x2, data = knn_data, xlim = c(-2, 2), ylim = c(-2, 10), pch = 20, col = orange_blue)
points(1.7, 10)
plot(scale(x1) ~ scale(x2), data = knn_data, xlim = c(-2, 2), ylim = c(-2, 10), pch = 20, col = orange_blue)
points((1.7 - 1.197685) / 0.6072974, (10 - 5) / 2.974975)
```


## Curse of Dimensionality

```{r}
set.seed(42)
knn_data_trn = sim_knn_data()
knn_data_tst = sim_knn_data()
```


## `rmarkdown`

The `rmarkdown` file for this chapter can be found [**here**](07-knn-reg.Rmd). The file was created using `R` version `r paste0(version$major, "." ,version$minor)`. The following packages (and their dependencies) were loaded when knitting this file:

```{r, echo = FALSE}
names(sessionInfo()$otherPkgs)
```
