# Logistic Regression

In this chapter, we continue our discussion of classification. We introduce our first model for classification, logistic regression. To begin, we return to the `Default` dataset from the previous chapter. 

```{r}
library(ISLR)
library(tibble)
as_tibble(Default)
```

We also repeat the test-train split from the previous chapter.

```{r}
set.seed(42)
default_index = sample(nrow(Default), 5000)
default_train = Default[default_index, ]
default_test = Default[-default_index, ]
```


## Linear Regression

Before moving on to logistic regression, why not plain, old, linear regression?

```{r}
default_train_lm = default_train
default_test_lm = default_test
```

Since linear regression expects a numeric response variable, we coerce the response to be numeric. (Notice that we also shift the results, as we require `0` and `1`, not `1` and `2`.) Notice we have also copied the dataset so that we can return the original data with factors later.

```{r}
default_train_lm$default = as.numeric(default_train_lm$default) - 1
default_test_lm$default = as.numeric(default_test_lm$default) - 1
```

Why would we think this should work? Recall that,

$$
\hat{E}[Y \mid X = x] = X\hat{\beta}.
$$

Since $Y$ is limited to values of $0$ and $1$, we have

$$
E[Y \mid X = x] = P[Y = 1 \mid X = x].
$$

It would then seem reasonable that $X\hat{\beta}$ is a reasonable estimate of $P[Y = 1 \ X = x]$. We test this on the `Default` data.

```{r}
model_lm = lm(default ~ balance, data = default_train_lm)
```

Everything seems to be working, until we plot the results.

```{r}
plot(default ~ balance, data = default_train_lm, 
     col = "darkorange", pch = "|", ylim = c(-0.2, 1),
     main = "Using Linear Regression for Classification")
abline(h = 0, lty = 3)
abline(h = 1, lty = 3)
abline(h = 0.5, lty = 2)
abline(model_lm, lwd = 3, col = "dodgerblue")
```

Two issues arise. First, all of the predicted probabilities are below 0.5 That means, we would classify every observation as a `"No"`. This is certainly possible, but not what we would expect.

```{r}
all(predict(model_lm) < 0.5)
```

The next, and bigger issue, is predicted probabilities less than 0.

```{r}
any(predict(model_lm) < 0)
```

## Bayes Classifier

Why are we using a predicted probability of 0.5 as the cutoff for classification? Recall, the Bayes Classifier, which minimizes the classification error:

$$
C^B({\bf x}) = \underset{k}{\mathrm{argmax}} \ P[Y = k \mid {\bf X = x}]
$$

So, in the binary classification problem, we will use predicted probabilities

$$
\hat{p}({\bf x}) = \hat{P}[Y = 1 \mid {\bf X = x}]
$$

and 

$$
\hat{P}[Y = 0 \mid {\bf X = x}]
$$

and then classify to the larger of the two. We actually only need to consider a single probability, usually for $\hat{P}[Y = 1 \mid {\bf X = x}]$. Since we use it so often, we give it the shorthand notation, $\hat{p}({\bf x})$. Then the classifier is written,

$$
\hat{C}(\bf x) = 
\begin{cases} 
      1 & \hat{p}({\bf x}) > 0.5 \\
      0 & \hat{p}({\bf x}) \leq 0.5 
\end{cases}
$$

## Logistic Regression with `glm()`

To better estimate the probability

$$
p({\bf x}) = P[Y = 1 \mid {\bf X = x}]
$$
we turn to logistic regression. The model is written

$$
\log\left(\frac{p({\bf x})}{1 - p({\bf x})}\right) = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \cdots  + \beta_p x_p.
$$

Rearranging, we see the probabilities can be written as

$$
p({\bf x}) = \frac{1}{1 + e^{-(\beta_0 + \beta_1 x_1 + \beta_2 x_2 + \cdots  + \beta_p x_p)}} = \sigma(\beta_0 + \beta_1 x_1 + \beta_2 x_2 + \cdots  + \beta_p x_p)
$$

Notice, we use the sigmoid function as shorthand notation, which appears often in deep learning literature. It takes any real input, and outputs a number between 0 and 1. How useful!

$$
\sigma(x) = \frac{1}{1 + e^{-x}}
$$

The model is fit by numerically maximizing the likelihood, which we will let `R` take care of.

We start with a single predictor example, again using `balance` as our single predictor.

```{r}
model_glm = glm(default ~ balance, data = default_train, family = "binomial")
```

Fitting this model looks very similar to fitting a simple linear regression. Instead of `lm()` we use `glm()`. The only other difference is the use of `family = "binomial"` which indicates that we have a two-class categorical response. Using `glm()` with `family = "gaussian"` would perform the usual linear regression.

First, we can obtain the fitted coefficients the same way we did with linear regression.

```{r}
coef(model_glm)
```

The next thing we should understand is how the `predict()` function works with `glm()`. So, let's look a some predictions.

```{r}
head(predict(model_glm))
```

By default, `predict.glm()` uses `type = "link"`.

```{r}
head(predict(model_glm, type = "link"))
```

That is, `R` is returning

$$
\hat{\beta}_0 + \hat{\beta}_1 x_1 + \hat{\beta}_2 x_2 + \cdots  + \hat{\beta}_p x_p
$$
for each observation.

Importantly, these are **not** predicted probabilities. To obtain the predicted probabilities

$$
\hat{p}({\bf x}) = \hat{P}[Y = 1 \mid {\bf X = x}]
$$

we need to use `type = "response"`

```{r}
head(predict(model_glm, type = "response"))
```

Note that these are probabilities, **not** classifications. To obtain classifications, we will need to compare to the correct cutoff value with an `ifelse()` statement.

```{r}
model_glm_pred = ifelse(predict(model_glm, type = "link") > 0, "Yes", "No")
# model_glm_pred = ifelse(predict(model_glm, type = "response") > 0.5, "Yes", "No")
```

The line that is run is performing

$$
\hat{C}(\bf x) = 
\begin{cases} 
      1 & \hat{f}({\bf x}) > 0.5 \\
      0 & \hat{f}({\bf x}) \leq 0.5 
\end{cases}
$$

where 

$$
\hat{f}({\bf x}) =\hat{\beta}_0 + \hat{\beta}_1 x_1 + \hat{\beta}_2 x_2 + \cdots  + \hat{\beta}_p x_p.
$$

The commented line, which would give the same results, is performing


$$
\hat{C}(\bf x) = 
\begin{cases} 
      1 & \hat{p}({\bf x}) > 0.5 \\
      0 & \hat{p}({\bf x}) \leq 0.5 
\end{cases}
$$

where 

$$
\hat{p}({\bf x}) = \hat{P}[Y = 1 \mid {\bf X = x}].
$$
Once we have classifications, we can calculate metrics such as accuracy.

```{r}
mean(model_glm_pred == default_train$default) # train accuracy
```

As we saw previously, the `table()` and `confusionMatrix()` functions can be used to quickly obtain many more metrics.

```{r, message = FALSE, warning = FALSE}
train_tab = table(predicted = model_glm_pred, actual = default_train$default)
library(caret)
train_con_mat = confusionMatrix(train_tab, positive = "Yes")
c(train_con_mat$overall["Accuracy"], 
  train_con_mat$byClass["Sensitivity"], 
  train_con_mat$byClass["Specificity"])
```

As we did with regression, we could also write a custom function for accuracy.

```{r}
get_accuracy = function(mod, data, res = "y", pos = 1, neg = 0, cut = 0.5) {
  probs = predict(mod, newdata = data, type = "response")
  preds = ifelse(probs > cut, pos, neg)
  mean(data[, res] == preds)
}
```

This function will be useful later when calculating train and test accuracies for several models at the same time.

```{r}
get_accuracy(model_glm, data = default_train, 
             res = "default", pos = "Yes", neg = "No", cut = 0.5)
```

To see how much better logistic regression is for this task, we create the same plot we used for linear regression.

```{r}
plot(default ~ balance, data = default_train_lm, 
     col = "darkorange", pch = "|", ylim = c(-0.2, 1),
     main = "Using Logistic Regression for Classification")
abline(h = 0, lty = 3)
abline(h = 1, lty = 3)
abline(h = 0.5, lty = 2)
curve(predict(model_glm, data.frame(balance = x), type = "response"), 
      add = TRUE, lwd = 3, col = "dodgerblue")
abline(v = -coef(model_glm)[1] / coef(model_glm)[2], lwd = 2)
```

This plot contains a wealth of information.

- The orange `|` characters are the data, $(x_i, y_i)$.
- The blue "curve" is the predicted probabilities given by the fitted logistic regression. That is,
$$
\hat{p}({\bf x}) = \hat{P}[Y = 1 \mid {\bf X = x}]
$$
- The solid vertical black line represents the **[decision boundary](https://en.wikipedia.org/wiki/Decision_boundary)**, the `balance` that obtains a predicted probability of 0.5. In this case `balance` = `r -coef(model_glm)[1] / coef(model_glm)[2]`.

The decision boundary is found by solving for points that satisfy

$$
\hat{p}({\bf x}) = \hat{P}[Y = 1 \mid {\bf X = x}] = 0.5
$$

This is equivalent to point that satisfy

$$
\hat{\beta}_0 + \hat{\beta}_1 x_1 = 0.
$$
Thus, for logistic regression with a single predictor, the decision boundary is given by the *point*

$$
x_1 = \frac{-\hat{\beta}_0}{\hat{\beta}_1}.
$$

The following is not run, but an alternative way to add the logistic curve to the plot.

```{r, eval = FALSE}
grid = seq(0, max(default_train$balance), by = 0.01)

sigmoid = function(x) {
  1 / (1 + exp(-x))
}

lines(grid, sigmoid(coef(model_glm)[1] + coef(model_glm)[2] * grid), lwd = 3)
```


Using the usual formula syntax, it is easy to add complexity to logistic regressions.

```{r}
model_1 = glm(default ~ 1, data = default_train, family = "binomial")
model_2 = glm(default ~ ., data = default_train, family = "binomial")
model_3 = glm(default ~ . ^ 2 + I(balance ^ 2),
              data = default_train, family = "binomial")
```

Note that, using polynomial transformations of predictors will allow a linear model to have non-linear decision boundaries.

```{r}
model_list = list(model_1, model_2, model_3)

train_error = 1 - sapply(model_list, get_accuracy, data = default_train, 
                         res = "default", pos = "Yes", neg = "No", cut = 0.5)
test_error = 1 - sapply(model_list, get_accuracy, data = default_test, 
                       res = "default", pos = "Yes", neg = "No", cut = 0.5)
```

Here we see the misclassification error rates for each model. The train decreases, and the test decreases, until it starts to increases. Everything we learned about the bias-variance tradeoff for regression also applies here.

```{r}
diff(train_error)
diff(test_error)

```

We call `model_2` the **additive** logistic model, which we will use quite often.


## ROC Curves

Let's return to our simple model with only balance as a predictor.


```{r}
model_glm = glm(default ~ balance, data = default_train, family = "binomial")
```

We write a function which allows use to make predictions based on different probability cutoffs.

```{r}
get_pred = function(mod, data, res = "y", pos = 1, neg = 0, cut = 0.5) {
  probs = predict(mod, newdata = data, type = "response")
  ifelse(probs > cut, pos, neg)
}
```


$$
\hat{C}(\bf x) = 
\begin{cases} 
      1 & \hat{f}({\bf x}) > c \\
      0 & \hat{f}({\bf x}) \leq c 
\end{cases}
$$

Let's use this to obtain predictions using a low, medium, and high cutoff. (0.1, 0.5, and 0.9)

```{r}
test_pred_10 = get_pred(model_glm, data = default_test, res = "default", pos = "Yes", neg = "No", cut = 0.1)
test_pred_50 = get_pred(model_glm, data = default_test, res = "default", pos = "Yes", neg = "No", cut = 0.5)
test_pred_90 = get_pred(model_glm, data = default_test, res = "default", pos = "Yes", neg = "No", cut = 0.9)
```

Now we evaluate accuracy, sensitivity, and specificity for these classifiers.

```{r}
test_tab_10 = table(predicted = test_pred_10, actual = default_train$default)
test_tab_50 = table(predicted = test_pred_50, actual = default_train$default)
test_tab_90 = table(predicted = test_pred_90, actual = default_train$default)


test_con_mat_10 = confusionMatrix(test_tab_10, positive = "Yes")
test_con_mat_50 = confusionMatrix(test_tab_50, positive = "Yes")
test_con_mat_90 = confusionMatrix(test_tab_90, positive = "Yes")
```

```{r}
metrics = rbind(
  
  c(test_con_mat_10$overall["Accuracy"], 
    test_con_mat_10$byClass["Sensitivity"], 
    test_con_mat_10$byClass["Specificity"]),
  
  c(test_con_mat_50$overall["Accuracy"], 
    test_con_mat_50$byClass["Sensitivity"], 
    test_con_mat_50$byClass["Specificity"]),
  
  c(test_con_mat_90$overall["Accuracy"], 
    test_con_mat_90$byClass["Sensitivity"], 
    test_con_mat_90$byClass["Specificity"])
  
)

rownames(metrics) = c("c = 0.10", "c = 0.50", "c = 0.90")
metrics
```

We see then sensitivity decreases as the cutoff is increases. Conversely, specificity increases as the cutoff increases. This is useful if we are more interested in a particular error, instead of giving them equal weight.

Note that usually the best accuracy will be seen near $c = 0.50$, but our results here difference due to the randomness of the test-train split.

Instead of manually checking cutoffs, we can create an ROC curve (receiver operating characteristic curve) which will sweep through all possible cutoffs, and plot the sensitivity and specificity.

```{r, message = FALSE, warning = FALSE}
library(pROC)
test_prob = predict(model_glm, newdata = default_test, type = "response")
test_roc = roc(default_test$default ~ test_prob, plot = TRUE, print.auc = TRUE)
as.numeric(test_roc$auc)
```

A good model will have a high AUC, that is as often as possible a hgih sensitivity and specificity.

## Multinomial Logistic Regression

What if the response contains more than two categories? For that we need multinomial logistic regression. 

$$
P[Y = k \mid {\bf X = x}] = \frac{e^{\beta_{0k} + \beta_{1k} x_1 + \cdots +  + \beta_{pk} x_p}}{\sum_{j = 1}^{K} e^{\beta_{0j} + \beta_{1j} x_1 + \cdots +  + \beta_{pj} x_p}}
$$

We will omit the details, as ISL has as well. If you are interested, the [Wikipedia page](https://en.wikipedia.org/wiki/Multinomial_logistic_regression) provides a rather thorough coverage. Also note that the above is an example of the [softmax function](https://en.wikipedia.org/wiki/Softmax_function).

As an example of a dataset with a three category response, we use the `iris` dataset, which is so famous, it has its own [Wikipedia entry](https://en.wikipedia.org/wiki/Iris_flower_data_set). It is also a default dataset in `R`, so no need to load it.

Before proceeding, we test-train split this data.

```{r}
iris_obs = nrow(iris)
iris_index = sample(iris_obs, size = trunc(0.50 * iris_obs))
iris_train = iris[iris_index, ]
iris_test = iris[-iris_index, ]
```

To perform multinomial logistic regression, we use the `multinom` function from the `nnet` package. Training using `multinom()` is done using similar syntax to `lm()` and `glm()`. We add the `trace = FALSE` argument to suppress information about updates to the optimization routine as the model is trained.

```{r}
library(nnet)
model_multi = multinom(Species ~ ., data = iris_train, trace = FALSE)
summary(model_multi)$coefficients
```

Notice we are only given coefficients for two of the three class, much like only needing coefficients for one class in logistic regression.

A difference between `glm()` and `multinom()` is how the `predict()` function operates.

```{r}
head(predict(model_multi, newdata = iris_train))
head(predict(model_multi, newdata = iris, type = "prob"))
```

Notice that by default, classifications are returned. When obtaining probabilities, we are given the predicted probability for **each** class.

Interestingly, you've just fit a neural network, and you didn't even know it! (Hence the `nnet` package.) Later we will discuss the connections between logistic regression, multinomial logistic regression, and simple neural networks.
