# Trees

**Chapter Status:** This chapter was originally written using the `tree` packages. Currently being re-written to exclusively use the `rpart` package which seems more widely suggested and provides better plotting features.

```{r tree_opts, include = FALSE}
knitr::opts_chunk$set(cache = TRUE, autodep = TRUE, fig.align = "center")
```

```{r, message = FALSE, warning = FALSE}
library(tree)
```

In this document, we will use the package `tree` for both classification and regression trees. Note that there are many packages to do this in `R`. [`rpart`](https://cran.r-project.org/web/packages/rpart/vignettes/longintro.pdf) may be the most common, however, we will use `tree` for simplicity.


## Classification Trees

```{r}
library(ISLR)
```

To understand classification trees, we will use the `Carseat` dataset from the `ISLR` package. We will first modify the response variable `Sales` from its original use as a numerical variable, to a categorical variable with `High` for high sales, and `Low` for low sales.

```{r}
data(Carseats)
#?Carseats
str(Carseats)
Carseats$Sales = as.factor(ifelse(Carseats$Sales <= 8, "Low", "High"))
str(Carseats)
```

We first fit an unpruned classification tree using all of the predictors. Details of this process can be found using `?tree` and `?tree.control`

```{r}
seat_tree = tree(Sales ~ ., data = Carseats)
# seat_tree = tree(Sales ~ ., data = Carseats,
#                  control = tree.control(nobs = nrow(Carseats), minsize = 10))
summary(seat_tree)
```

We see this tree has 27 terminal nodes and a misclassification rate of 0.09.

```{r, fig.height = 12, fig.width = 24}
plot(seat_tree)
text(seat_tree, pretty = 0)
title(main = "Unpruned Classification Tree")
```

Above we plot the tree. Below we output the details of the splits.

```{r}
seat_tree
```

We now test-train split the data so we can evaluate how well our tree is working. We use 200 observations for each.

```{r}
dim(Carseats)
set.seed(2)
seat_idx = sample(1:nrow(Carseats), 200)
seat_trn = Carseats[seat_idx,]
seat_tst = Carseats[-seat_idx,]
```

```{r}
seat_tree = tree(Sales ~ ., data = seat_trn)
```

```{r}
summary(seat_tree)
```

Note that, the tree is not using all of the available variables.

```{r}
summary(seat_tree)$used
names(Carseats)[which(!(names(Carseats) %in% summary(seat_tree)$used))]
```

Also notice that, this new tree is slightly different than the tree fit to all of the data.

```{r, fig.height=12, fig.width=24}
plot(seat_tree)
text(seat_tree, pretty = 0)
title(main = "Unpruned Classification Tree")
```

When using the `predict()` function on a tree, the default `type` is `vector` which gives predicted probabilities for both classes. We will use `type = class` to directly obtain classes. We first fit the tree using the training data (above), then obtain predictions on both the train and test set, then view the confusion matrix for both.

```{r}
seat_trn_pred = predict(seat_tree, seat_trn, type = "class")
seat_tst_pred = predict(seat_tree, seat_tst, type = "class")
#predict(seat_tree, seat_trn, type = "vector")
#predict(seat_tree, seat_tst, type = "vector")
```


```{r}
# train confusion
table(predicted = seat_trn_pred, actual = seat_trn$Sales)
# test confusion
table(predicted = seat_tst_pred, actual = seat_tst$Sales)
```

```{r}
accuracy = function(actual, predicted) {
  mean(actual == predicted)
}
```


```{r}
# train acc
accuracy(predicted = seat_trn_pred, actual = seat_trn$Sales)
# test acc
accuracy(predicted = seat_tst_pred, actual = seat_tst$Sales)
```

Here it is easy to see that the tree has been over-fit. The train set performs much better than the test set.

We will now use cross-validation to find a tree by considering trees of different sizes which have been pruned from our original tree.

```{r}
set.seed(3)
seat_tree_cv = cv.tree(seat_tree, FUN = prune.misclass)
```

```{r}
# index of tree with minimum error
min_idx = which.min(seat_tree_cv$dev)
min_idx
```

```{r}
# number of terminal nodes in that tree
seat_tree_cv$size[min_idx]
```

```{r}
# misclassification rate of each tree
seat_tree_cv$dev / length(seat_idx)
```


```{r}
par(mfrow = c(1, 2))
# default plot
plot(seat_tree_cv)
# better plot
plot(seat_tree_cv$size, seat_tree_cv$dev / nrow(seat_trn), type = "b",
     xlab = "Tree Size", ylab = "CV Misclassification Rate")
```

It appears that a tree of size 9 has the fewest misclassifications of the considered trees, via cross-validation.

We use `prune.misclass()` to obtain that tree from our original tree, and plot this smaller tree.

```{r}
seat_tree_prune = prune.misclass(seat_tree, best = 9)
summary(seat_tree_prune)
```

```{r, fig.height=8, fig.width=12}
plot(seat_tree_prune)
text(seat_tree_prune, pretty = 0)
title(main = "Pruned Classification Tree")
```

We again obtain predictions using this smaller tree, and evaluate on the test and train sets.

```{r}
# train
seat_prune_trn_pred = predict(seat_tree_prune, seat_trn, type = "class")
table(predicted = seat_prune_trn_pred, actual = seat_trn$Sales)
accuracy(predicted = seat_prune_trn_pred, actual = seat_trn$Sales)
```

```{r}
# test
seat_prune_tst_pred = predict(seat_tree_prune, seat_tst, type = "class")
table(predicted = seat_prune_tst_pred, actual = seat_tst$Sales)
accuracy(predicted = seat_prune_tst_pred, actual = seat_tst$Sales)
```

The train set has performed almost as well as before, and there was a **small** improvement in the test set, but it is still obvious that we have over-fit. Trees tend to do this. We will look at several ways to fix this, including: bagging, boosting and random forests.


## Regression Trees

To demonstrate regression trees, we will use the `Boston` data. Recall `medv` is the response. We first split the data in half.

```{r}
library(MASS)
set.seed(18)
boston_idx = sample(1:nrow(Boston), nrow(Boston) / 2)
boston_trn = Boston[boston_idx,]
boston_tst = Boston[-boston_idx,]
```

Then fit an unpruned regression tree to the training data.

```{r}
boston_tree = tree(medv ~ ., data = boston_trn)
summary(boston_tree)
```

```{r, fig.height=8, fig.width=12}
plot(boston_tree)
text(boston_tree, pretty = 0)
title(main = "Unpruned Regression Tree")
```

As with classification trees, we can use cross-validation to select a good pruning of the tree.

```{r}
set.seed(18)
boston_tree_cv = cv.tree(boston_tree)
plot(boston_tree_cv$size, sqrt(boston_tree_cv$dev / nrow(boston_trn)), type = "b",
     xlab = "Tree Size", ylab = "CV-RMSE")
```

While the tree of size 9 does have the lowest RMSE, we'll prune to a size of 7 as it seems to perform just as well. (Otherwise we would not be pruning.) The pruned tree is, as expected, smaller and easier to interpret.

```{r}
boston_tree_prune = prune.tree(boston_tree, best = 7)
summary(boston_tree_prune)
```

```{r, fig.height=8, fig.width=12}
plot(boston_tree_prune)
text(boston_tree_prune, pretty = 0)
title(main = "Pruned Regression Tree")
```

Let's compare this regression tree to an additive linear model and use RMSE as our metric.

```{r}
rmse = function(actual, predicted) {
  sqrt(mean((actual - predicted) ^ 2))
}
```


We obtain predictions on the train and test sets from the pruned tree. We also plot actual vs predicted. This plot may look odd. We'll compare it to a plot for linear regression below.

```{r}
# training RMSE two ways
sqrt(summary(boston_tree_prune)$dev / nrow(boston_trn))
boston_prune_trn_pred = predict(boston_tree_prune, newdata = boston_trn)
rmse(boston_prune_trn_pred, boston_trn$medv)
```

```{r}
# test RMSE
boston_prune_tst_pred = predict(boston_tree_prune, newdata = boston_tst)
rmse(boston_prune_tst_pred, boston_tst$medv)
```


```{r}
plot(boston_prune_tst_pred, boston_tst$medv, xlab = "Predicted", ylab = "Actual")
abline(0, 1)
```

Here, using an additive linear regression the actual vs predicted looks much more like what we are used to.

```{r}
bostom_lm = lm(medv ~ ., data = boston_trn)
boston_lm_pred = predict(bostom_lm, newdata = boston_tst)
plot(boston_lm_pred, boston_tst$medv, xlab = "Predicted", ylab = "Actual")
abline(0, 1)
rmse(boston_lm_pred, boston_tst$medv)
```

We also see a lower test RMSE. The most obvious linear regression beats the tree! Again, we'll improve on this tree soon. Also note the summary of the additive linear regression below. Which is easier to interpret, that output, or the small tree above?

```{r}
coef(bostom_lm)
```


## `rpart` Package

The `rpart` package is an alternative method for fitting trees in `R`. It is much more feature rich, including fitting multiple cost complexities and performing cross-validation by default. It also has the ability to produce much nicer trees. Based on its default settings, it will often result in smaller trees than using the `tree` package. See the references below for more information. `rpart` can also be tuned via `caret`.

```{r}
library(rpart)
set.seed(430)
# Fit a decision tree using rpart
# Note: when you fit a tree using rpart, the fitting routine automatically
# performs 10-fold CV and stores the errors for later use
# (such as for pruning the tree)

# fit a tree using rpart
seat_rpart = rpart(Sales ~ ., data = seat_trn, method = "class")

# plot the cv error curve for the tree
# rpart tries different cost-complexities by default
# also stores cv results
plotcp(seat_rpart)

# find best value of cp
min_cp = seat_rpart$cptable[which.min(seat_rpart$cptable[,"xerror"]),"CP"]
min_cp

# prune tree using best cp
seat_rpart_prune = prune(seat_rpart, cp = min_cp)

# nicer plots
library(rpart.plot)
prp(seat_rpart_prune)
prp(seat_rpart_prune, type = 4)
rpart.plot(seat_rpart_prune)
```

## External Links

- [An Introduction to Recursive Partitioning Using the `rpart` Routines](https://cran.r-project.org/web/packages/rpart/vignettes/longintro.pdf) - Details of the `rpart` package.
- [`rpart.plot` Package](http://www.milbo.org/doc/prp.pdf) - Detailed manual on plotting with `rpart` using the `rpart.plot` package.


## `rmarkdown`

The `rmarkdown` file for this chapter can be found [**here**](26-tree.Rmd). The file was created using `R` version `r paste0(version$major, "." ,version$minor)`. The following packages (and their dependencies) were loaded when knitting this file:

```{r, echo = FALSE}
names(sessionInfo()$otherPkgs)
```
