# Simulating the Bias–Variance Tradeoff

## Bias-Variance Decomposition

$$
y = f(\mathbf x) + \epsilon
$$


- estimate $\hat{f}(\mathbf x)$, train using data

$$
E[\epsilon] = 0 \quad \text{and} \quad \text{var}(\epsilon) = \sigma^2.
$$

- expected prediction error at $x_0$

$$
E\left[\left(y_0 - \hat{f}(\mathbf x_0)\right)^2\right] = \text{bias}\left(\hat{f}(\mathbf x_0)\right)^2 + \text{var}\left(\hat{f}(\mathbf x_0)\right) + \sigma^2
$$



$$
\text{bias}\left(\hat{f}(\mathbf x_0)\right) = E\left[\hat{f}(\mathbf x_0)\right] - f(\mathbf x_0)
$$

- reducible error = bias^2 = var
- irreduble error, noise


## Simulation

Suppose we would like a train a model to learn the function $f(x) = x^2$.

```{r}
f = function(x) {
  x ^ 2
}
```

More specifically,

$$
y = x^2 + \epsilon
$$
where

$$
\epsilon \sim N(\mu = 0, \sigma^2 = 0.3^2).
$$

We write a function which generate data accordingly.

```{r}
get_sim_data = function(f, sample_size = 100) {
  x = runif(n = sample_size, 0, 1)
  y = f(x) + rnorm(n = sample_size, mean = 0, sd = 0.3)
  data.frame(x, y)
}
```

- TODO: fitting to some data

```{r}
sim_data = get_sim_data(f, sample_size = 100)

fit_1 = lm(y ~ 1, data = sim_data)
fit_2 = lm(y ~ poly(x, degree = 1), data = sim_data)
fit_3 = lm(y ~ poly(x, degree = 2), data = sim_data)
fit_4 = lm(y ~ poly(x, degree = 3), data = sim_data)
```

- TODO: plotting data and fits

```{r, fig.height = 8, fig.width = 8}
plot(y ~ x, data = sim_data)
grid = seq(from = 0, to = 1, by = 0.01)
lines(grid, f(grid), col = "black", lwd = 5)
lines(grid, predict(fit_1, newdata = data.frame(x = grid)), col = "red", lwd = 2)
lines(grid, predict(fit_2, newdata = data.frame(x = grid)), col = "blue", lwd = 2)
lines(grid, predict(fit_3, newdata = data.frame(x = grid)), col = "green", lwd = 3)
lines(grid, predict(fit_4, newdata = data.frame(x = grid)), col = "orange", lwd = 2)
```

- TODO: repeated simulations

```{r, fig.height = 10, fig.width = 10}
set.seed(1)
n_sims = 1000
n_models = 4
x0 = 0.9
predictions = matrix(0, nrow = n_sims, ncol = n_models)
sim_data = get_sim_data(f, sample_size = 100)
plot(y ~ x, data = sim_data, col = "white", xlim = c(0.75, 1), ylim = c(0, 1.5))

for (i in 1:n_sims) {
  
  sim_data = get_sim_data(f, sample_size = 100)

  fit_1 = lm(y ~ 1, data = sim_data)
  fit_2 = lm(y ~ poly(x, degree = 1), data = sim_data)
  fit_3 = lm(y ~ poly(x, degree = 2), data = sim_data)
  fit_4 = lm(y ~ poly(x, degree = 3), data = sim_data)
  
  lines(grid, predict(fit_1, newdata = data.frame(x = grid)), col = "red", lwd = 1)
  lines(grid, predict(fit_2, newdata = data.frame(x = grid)), col = "blue", lwd = 1)
  lines(grid, predict(fit_3, newdata = data.frame(x = grid)), col = "green", lwd = 1)
  lines(grid, predict(fit_4, newdata = data.frame(x = grid)), col = "orange", lwd = 1)

  predictions[i, ] <- c(
    predict(fit_1, newdata = data.frame(x = x0)),
    predict(fit_2, newdata = data.frame(x = x0)),
    predict(fit_3, newdata = data.frame(x = x0)),
    predict(fit_4, newdata = data.frame(x = x0))
  )
}

lines(grid, f(grid), col = "white", lwd = 5)
lines(grid, f(grid), col = "black", lwd = 5, lty = 2)
```

- TODO: new obs at particular point x0

```{r}
eps = rnorm(n = n_sims, mean = 0, sd = 0.3)
y0 = f(x0) + eps
```


## Bias-Variance Tradeoff

```{r}
get_bias = function(estimate, truth) {
  mean(estimate - truth)
}

get_mse = function(estimate, truth) {
  mean((estimate - truth) ^ 2)
}

```

```{r}
bias = apply(predictions, 2, get_bias, y0)
variance = apply(predictions, 2, var)
mse = apply(predictions, 2, get_mse, y0)
```

```{r}
bias ^ 2 + variance + var(eps)
mse
```

```{r, echo = FALSE}
options(scipen = 999)
```

| Model   | Squared Bias     | Variance (Of Estimate) | MSE        |
|---------|------------------|------------------------|------------|
| `fit_1` | `r bias[1] ^ 2`  | `r variance[1]`        | `r mse[1]` |
| `fit_2` | `r bias[2] ^ 2`  | `r variance[2]`        | `r mse[2]` |
| `fit_3` | `r bias[3] ^ 2`  | `r variance[3]`        | `r mse[3]` |
| `fit_4` | `r bias[4] ^ 2`  | `r variance[4]`        | `r mse[4]` |


```{r, echo = FALSE}
options(scipen = 0)
```


```{r}
all(diff(bias ^ 2) < 0)
all(diff(variance) > 0)
diff(mse)
```


- TODO: note about difference do to estimating with simulation


- TODO: summarize with table, talk about where balance is found, bias inc, var dec
