# The `caret` Package

Now that we have seen a number of classificaiton (and regression) methods, and introduced cross-validation, we see the general outline of a predictive analysis:

- Select a method
- Test-train split the data
- Decide on a set of candidate models via a tuning parameter
- Select best model (tuning parameters) via cross-validation
- Use chosen model to make predictions
- Calculate relevent metrics on the test data

At face value it would seem like it should be easy to repeat this process for a number of different methods, however we have run into a number of difficulties attempting to do so with `R`.

- The `predict()` function seems to have a different behvaior for each new method we see.
- Many methods have different cross-validation functions, or worse yet, no build-in process for cross-validation.
- Different methods have different handling of categorical predictors.

Thankfully, the `R` community has essentailly provided a silver bullet for these issues, the [`caret`](http://topepo.github.io/caret/) package.


- `train()`
    - formula
    - data
    - method
    
- `predict()`



- `createDataPartition()`
- `trainControl()`


- `expand.grid()`

- preProcess()


```{r}
data(Default, package = "ISLR")
```

```{r, message = FALSE, warning = FALSE}
library(caret)
```


```{r}
set.seed(430)
default_idx = createDataPartition(Default$default, p = 0.75, list = FALSE)
default_trn = Default[default_idx, ]
default_tst = Default[-default_idx, ]
```

```{r}
default_glm = train(
  default ~ .,
  data = default_trn,
  method = "glm",
  family = "binomial",
  trControl = trainControl(method = "cv", number = 5)
)
```

```{r}
default_glm
```


```{r}
names(default_glm)
```



```{r}
default_glm$results
```


```{r}
default_glm$finalModel
```

```{r}
accuracy = function(actual, predicted) {
  mean(actual == predicted)
}
```


```{r}
# make preds

head(predict(default_glm, newdata = default_trn))
```



```{r}
# train acc
accuracy(actual = default_trn$default,
         predicted = predict(default_glm, newdata = default_trn))
```

```{r}
# test acc
accuracy(actual = default_tst$default,
         predicted = predict(default_glm, newdata = default_tst))
```



```{r}
# get probs
head(predict(default_glm, newdata = default_trn, type = "prob"))
```















```{r}
default_knn = train(
  default ~ .,
  data = default_trn,
  method = "knn",
  trControl = trainControl(method = "cv", number = 5)
)
```





```{r}
default_knn
```





```{r}

```


```{r}
default_grid = expand.grid(k = seq(1, 151, by = 5))
```

- TODO: returns a data frame, a habit for when we later need multiple tuning parameters

```{r}
default_knn = train(
  default ~ .,
  data = default_trn,
  method = "knn",
  trControl = trainControl(method = "cv", number = 5, selectionFunction = "oneSE"),
  preProcess = c("center", "scale"),
  tuneGrid = default_grid
)
```



```{r}
default_knn
```



```{r}
plot(default_knn)
```



```{r}
ggplot(default_knn) + theme_bw()
```






```{r}
default_knn$bestTune
```


```{r}
get_best_result = function(caret_fit) {
  best_result = caret_fit$results[as.numeric(rownames(caret_fit$bestTune)), ]
  rownames(best_result) = NULL
  best_result
}
```

```{r}
get_best_result(default_knn)
```



```{r}
default_knn$finalModel
```










less ties with CV

run knn without a grid, notice default grid, also notice preprocess


TODO: table with best knn and glm results for train, test, cv



## External Links


[http://topepo.github.io/caret/index.html]

[http://topepo.github.io/caret/available-models.html]()


## RMarkdown

The RMarkdown file for this chapter can be found [**here**](13-caret.Rmd.Rmd). The file was created using `R` version `r paste0(version$major, "." ,version$minor)` and the following packages:

- Base Packages, Attached

```{r, echo = FALSE}
sessionInfo()$basePkgs
```

- Additonal Packages, Attached

```{r, echo = FALSE}
names(sessionInfo()$otherPkgs)
```

- Additonal Packages, Not Attached

```{r, echo = FALSE}
names(sessionInfo()$loadedOnly)
```




